{\plant}: A package for modelling forest trait ecology \& evolution

### 2015-05-28 ms.tex
More comments and changes

The extension for handling strong size-asymmetric competition involves
adaptive refining of time points at which new cohorts are introduced
into the population. In the original EBT, cohorts were introduced at a
fixed time interval. Under competition, the growth trajectories of
individuals born at similar times can diverge substantially over time
(Fig. \ref{fig:patch}a; Appendix \ref{sec:demography} and \ref{sec:cohort-spacing}).
This is can become a problem later during patch
development because much of the population can be located between
widely-spaced cohorts, delivering low numerical precision. To maintain
numerical precision we therefore use an iterative algorithm to adaptively refine
the cohort spacing until the trajectories of adjacent cohorts remain
adequately resolved throughout development of the patch (see
Appendix \ref{sec:demography} for details). This refinement causes an uneven
distribution of cohort introduction times across patch age (Fig.
\ref{fig:patch}a), with considerably tighter spacing of cohorts in 
younger patches.

@@

Most vegetation is subject to a disturbance regime, such that patches of
the landscape are cleared at some rate, by fire, cyclone, landslide, or
disease outbreak
\citep{Connell-1978, White-1979, Chambers-2013, Bormann-1979, Clark-1991, Coomes-2007}.
The vegetation thus comprises a collection of patches differing in time
since last disturbance and linked via seed dispersal (Fig.
\ref{fig:schematic}b). Such a system is often refereed to as a
structured metapopulation \citep{Gyllenberg-2001}.

Having simulated a single patch, it is straightforward to scale up from
a single patch to an entire metapopulation, at least for the situation
of metapopulations at an equilibrium mix of patch ages and where
disturbances remove all established vegetation. The frequency-density
\(p(a)\) of patches aged \(a\) in the landscape is needed. Assuming there
are a large number of patches means the dynamics of \(p\) behave
deterministically and can be approximated via a second PDE
\citep{Vonfoerster-1959, Mckendrick-1926} (see Appendix
\ref{sec:demography} for
details). Scaling from patch to metapopulation is then achieved by
weighting the temporal dynamics of a single patch by \(p(a)\), the
relative abundance of patches aged \(a\) in the metapopulation.

The main numerical challenge in stepping from single patch
to metapopulation is to identify the equilibrium seed rain of the
metapopulation. For a given seed rain input, the model produces a seed rain output. 
The equilibrium seed rain is the vector of seed rain
where input seed rain equals output seed rain. Finding this equilibrium
requires finding a root of a multidimensional function, subject to the
constraints that this root is stable (e.g., the trivial equilibrium of
zero seed rain always exists, but is often not stable) (Appendix \ref{sec:fitness});

@@

As previously noted \citep{Kohyama-1993, Moorcroft-2001, Falster-2011} a
feature of the size-structured metapopulation concept implemented here
is that it reconciles equilibrium and non-equilibrium approaches to
modelling ecological dynamics
\citep{Levin-1974, Bormann-1979, Connell-1978, Coomes-2007}. An
equilibrium may be approached at the level of metapopulation, meaning
the seed rain and size structure across the entire metapopulation is
approximately stable. Yet the structure of vegetation within individual
patches is constantly in flux. Patches continue to age and their
residents continue to grow until they are disturbed.

Beyond its conceptual appeal, this idea of a dynamic equilibrium allows
us to close the demographic feedback loop and thereby create a
self-regulating system. The equilibrium seed rain for multi-species
communities is approached demographically, incorporating the combined
effects of the model's physiological rules, species traits, competitive
interactions and the prevailing disturbance regime on the
size-distributions of plants within a metapopulation of patches. Unlike
some vegetation models, there is no need to set the density of seeds or
seedlings, these arise as outputs from the model. Characteristics of the
vegetation such as total leaf area, biomass, productivity and allocation
to different plant parts also arise as emergent properties (Fig.
\ref{fig:patch}c). Similarly, distributions of abundance, leaf area and
growth rate with respect to size emerge at both patch scale and scale of
metapopulation (Fig. \ref{fig:emergent}). Note in particular how a single
physiological model produces a variety of height-density relationships, 
dependent on patch age, as well as an emergent average across all patches 
(Fig. \ref{fig:emergent}). This emergent relationship might then be 
compared to the observed distributions in large forest plots \citep{Muller-2006}.

@@

With the extension to metapopulation, we are modelling demography across
the entire life cycle and are thus in a position to estimate the fitness
of different types. Invasion fitness quantifies the ability of a rare
individual with particular traits to establish in a community of
established residents at their equilibrium densities \citep{Metz-1992}.
Invasion fitness
is ideally calculated as the long-term per capita growth rate of the new
type; however, in structured metapopulations, a more convenient
indicator of per-capita growth rate is given by the logarithm of the
basic reproduction ratio \citep{Gyllenberg-2001, Metz-2001}. The basic
reproduction ratio \(R\) is simply the total seed output of the new
type, averaged across the metapopulation. The new type can invade when
\(R > 1\).

@@

Fitness landscapes will be positive in regions that allow invasion, and
zero for species that are at equilibrium. The slope of the fitness
landscape at a resident indicates the direction of selection; a negative
slope (dashed line in Fig. \ref{fig:fitness}a) indicates selection for
smaller trait values. Fitness landscapes are constantly reshaped
as species are introduced and evolve, reflecting the non-linear
demographic feedbacks that occur under frequency-dependent evolution
 \citep{Geritz-1998,Dieckmann-2007}. Following the fitness gradient, can lead to an
evolutionary branching point \citep{Geritz-1998,Dieckmann-2007};
at this point there is no directional selection (solid line in
figure \ref{fig:fitness}a) but invasion is possible both above and below
it in trait space. It is often possible
to find a configuration where fitness is negative everywhere (solid line
in Fig. \ref{fig:fitness}b) except for the zero points where species are
present. This community is immune to invasion from any other strategies
\citep{Geritz-1998}.

@@

{\plant} makes use of much existing software, including the programming
language C++; the R computational environment \citep{R-2015}, the R
packages \texttt{Rcpp} \citep{Eddelbuettel-2011, Eddelbuettel-2013},
\texttt{R6}\citep{Chang-2014}, and \texttt{testhat}
\citep{Wickham-2011}; and the Boost Library for C++
\citep{Schaling-2014} via the R package \texttt{BH}
\citep{Eddelbuettel-2015}. Source code is hosted at
\href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
Installation instructions are available at this homepage, and binary
releases will also be available there.

@@

The {\plant} package makes it easier for researchers and forest practitioners
alike to investigate a variety of non-linear demographic phenomena.  Rather
than provide a model with a few routines, we have developed an extensible
framework that we hope can be used to answer a number of questions at a
variety of scales. The specific use-cases highlighted here arise from our
interest in the assembly of trait mixtures in communities. However a variety
of other uses would be possible, and also extensions to address other facets
of community assembly. For example,  the physiology of growth could be
modified to account for rising atmospheric CO2 or the effects of competition
for multiple resources could likewise be investigated.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/plant.pdf}
\caption{Physiological dynamics for individual plants vary with size, trait
and light environment. (a) Growth trajectories for plants differing in light
environment and trait (LMA, low = 0.1, solid line; high =  0.267, dashed
line). Lower light environments slow trajectories and interact non-linearly
with traits. (b) Over time, the fraction of living tissue switches from leaf
towards sapwood, with high LMA species having relatively more mass in leaf
than low LMA species. (c) Growth rates peak around 5m high, but this position
varies with both trait and light level (note that this is the derivative of
panel (a)). (d) Growth rates vary with plant size (seedling and mature shown)
and with traits; growth rate differences are more pronounced in high light.
Whole plant light compensation points emerge from the model as the point where
growth rate is zero (x intercept).  The light level used in panels (a) and (c)
is indicated with dots.} [Could you put delta symbols on the y axis rather than d? Also there is alot going in these figures with 4 lines each. You could make it easier for the reader by adding the dashed lines to the keys]
\label{fig:plant}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/patch.pdf}
\caption{Population dynamics for two species competing within a patch. Lines
in panels (a) same growth trajectories for individuals on the
boundary of successive "cohorts" for each species, using a schedule of 
introduction times generated by an adaptive refinement algorithm.
 Red lines are the low LMA
species from figure \ref{fig:plant}, blue lines are the high LMA species.  The
initial growth rate advantage of the low LMA species (see Figure
\ref{fig:plant} c) means it quickly over tops the high LMA species,
suppressing its growth.  Self-thinning leaves space for the high LMA species
to establish below the canopy of the fast species. Panel (b) shows the same
trajectories as in panel (a), but with shading indicating the density $n$ of
individuals at given size and patch age (light = low density, dark = high density).  
(c) The leaf area index at ground
level for each species.  The dotted line is the total leaf area index (sum
across both species).}
\label{fig:patch}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/emergent.pdf}
\caption{Emergent size-distributions across a metapopulation.
Grey lines show patterns within individual patches differing in time since disturbance (age), with blue lines indicating patches of a specific age. Panels show (a) patterns of abundance, (b) leaf area and (c) height growth rate against size. Note that units of panels (a) and (b) are per unit height and per unit ground area, respectively.}
\label{fig:emergent}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/fitness.pdf}
\caption{Example fitness landscape for LMA, showing potential for stable
coexistence of multiple types.  Fitness is the log of the population growth
rate - at zero species are at equilibrium, above zero they will increase in
density.  (a) Landscape generated by a single species.  At LMA of 0.1 (the
dashed line) there is directional selection for smaller LMA.  At LMA of 0.08
(solid line) there is a branching point (see inset for more detail).  (b)
Landscape generated by two species, holing the first at LMA of 0.08.
Introducing a species at the maximum fitness value from panel (a) reduces the
region of positive fitness considerably (dashed line).  Dotted lines show
subsequent invasions and replacements of the second species, and the solid
line shows a stable fitness landscape where both species exist on a peak in
the fitness landscape.  At this point no further invasion is possible.}
\label{fig:fitness}
\end{figure}

@@

\section{Solving for equilibrium seed rain}\label{sec:seed}

@@

>>>>

The extension for handling strong size-asymmetric competition involves
adaptive refining of time points at which new cohorts are introduced
into the population. In the original EBT, cohorts were introduced at a
fixed time interval. Under competition, the growth trajectories of
individuals born at similar times can diverge substantially over time
(Fig. \ref{fig:patch}a; Appendix \ref{sec:demography},
\ref{sec:patch}, and \ref{sec:cohort-spacing}). This is can become a
problem later during patch development because much of the population
can be located between widely-spaced cohorts, delivering low numerical
precision. To maintain numerical precision we therefore use an
iterative algorithm to adaptively refine the cohort spacing until the
trajectories of adjacent cohorts remain adequately resolved throughout
development of the patch (see Appendix \ref{sec:demography} for
details). This refinement causes an uneven distribution of cohort
introduction times across patch age (Fig.  \ref{fig:patch}a), with
considerably tighter spacing of cohorts in younger patches.

@@

Most vegetation is subject to a disturbance regime, such that patches of
the landscape are cleared at some rate, by fire, cyclone, landslide, or
disease outbreak
% TODO: Too many references
\citep{Connell-1978, White-1979, Chambers-2013, Bormann-1979, Clark-1991, Coomes-2007}.
The vegetation thus comprises a collection of patches differing in time
since last disturbance and linked via seed dispersal (Fig.
\ref{fig:schematic}b). Such a system is often refereed to as a
structured metapopulation \citep{Gyllenberg-2001}.

% TODO[RGF]: Again, I'd drop the "design features" thing here - it's
% not helpful I think.
\subsection{Design features}

In {\plant} we assume an infinite number of patches all experiencing
the same disturbance regime, and asusme that the disturbances remove
all established vegetation.  With this approach we can scale up from
simulating a single patch to simulate an entire metapopulation.
% TODO: Is the 1926 paper *really* needed?
The frequency-density
\(p(a)\) of patches aged \(a\) in the landscape is needed. Assuming there
are a large number of patches means the dynamics of \(p\) behave
deterministically and can be approximated via a second PDE
\citep{Vonfoerster-1959, Mckendrick-1926} (see Appendix
\ref{sec:demography} for
details). Scaling from patch to metapopulation is then achieved by
weighting the temporal dynamics of a single patch by \(p(a)\), the
relative abundance of patches aged \(a\) in the metapopulation.

The main numerical challenge in stepping from single patch to
metapopulation is to identify the equilibrium seed rain of the
metapopulation. For a given seed rain input, the model produces a seed
rain output.
% TODO: need to be clearer about what seed rain output means in this
% context?
The equilibrium seed rain is the vector of seed rain where input seed
rain equals output seed rain. Finding this equilibrium requires
finding a root of a multidimensional function, subject to the
constraints that this root is stable (e.g., the trivial equilibrium of
zero seed rain always exists, but is often not stable) (Appendix
\ref{sec:equilibrium}).

@@

% TODO[RGF]: This section is closer to the design features than to the
% use cases, I think.  Not sure what to suggest here.  But we need
% something that indicates how this can be used.  I suggest a complete
% overhaul of the second paragraph as that gets closest to a way of
% using the model.

As previously noted \citep{Kohyama-1993, Moorcroft-2001, Falster-2011} a
feature of the size-structured metapopulation concept implemented here
is that it balances equilibrium and non-equilibrium approaches to
modelling ecological dynamics
\citep{Levin-1974, Bormann-1979, Connell-1978, Coomes-2007}. An
equilibrium may be approached at the level of metapopulation, meaning
the seed rain and size structure across the entire metapopulation is
approximately stable. Yet the structure of vegetation within individual
patches is constantly in flux. Patches continue to age and their
residents continue to grow until they are disturbed.

Beyond its conceptual appeal, this idea of a dynamic equilibrium
allows us to close the demographic feedback loop and thereby create a
self-regulating system. The equilibrium seed rain for multi-species
communities is approached demographically, incorporating the combined
effects of the model's physiological rules, species traits,
competitive interactions and the prevailing disturbance regime on the
size-distributions of plants within a metapopulation of
patches. Unlike some vegetation models, there is no need to set the
density of seeds or seedlings, these arise as outputs from the
model. Characteristics of the vegetation such as total leaf area,
biomass, productivity and allocation to different plant parts also
arise as emergent properties (Fig.  \ref{fig:patch}c). Similarly,
distributions of abundance, leaf area and growth rate with respect to
size emerge at both patch scale and scale of metapopulation
(Fig. \ref{fig:emergent}). Note in particular how a single
physiological model produces a variety of height-density
relationships, dependent on patch age, as well as an emergent average
across all patches (Fig. \ref{fig:emergent}, see more detail in
appendix \ref{sec:emergent}). This emergent relationship might then be
compared to the observed distributions in large forest plots
\citep{Muller-2006}.

@@

With the extension to metapopulation, we are modelling demography across
the entire life cycle and are thus in a position to estimate the fitness
of different types. Invasion fitness quantifies the ability of a rare
individual with particular traits to establish in a community of
established residents at their equilibrium densities \citep{Metz-1992}.
Invasion fitness
is ideally calculated as the long-term per capita growth rate of the new
type; however, in structured metapopulations, a more convenient
indicator of per-capita growth rate is given by the logarithm of the
basic reproduction ratio \citep{Gyllenberg-2001, Metz-2001}. The basic
reproduction ratio \(R\) is simply the total seed output of the new
type, averaged across the metapopulation. A new type can invade when
\(R > 1\).

@@

Fitness landscapes will be positive in regions that allow invasion,
and zero for species that are at equilibrium. The slope of the fitness
landscape at a resident indicates the direction of selection; a
negative slope (dashed line in Fig. \ref{fig:fitness}a) indicates
selection for smaller trait values (see appendix \ref{sec:fitness} for
details). Fitness landscapes are constantly reshaped as species are
introduced and evolve, reflecting the non-linear demographic feedbacks
that occur under frequency-dependent evolution
\citep{Geritz-1998,Dieckmann-2007}. Following the fitness gradient,
can lead to an evolutionary branching point
\citep{Geritz-1998,Dieckmann-2007}; at this point there is no
directional selection (solid line in figure \ref{fig:fitness}a) but
invasion is possible both above and below it in trait space. It is
often possible to find a configuration where fitness is negative
everywhere (solid line in Fig. \ref{fig:fitness}b) except for the zero
points where species are present. This community is immune to invasion
from any other strategies \citep{Geritz-1998}.

@@

{\plant} makes use of much existing software, including the programming
language C++; the R computational environment \citep{R-2015}, the R
packages \texttt{Rcpp} \citep{Eddelbuettel-2011, Eddelbuettel-2013},
\texttt{R6} \citep{Chang-2014}, and \texttt{testhat}
\citep{Wickham-2011}; and the Boost Library for C++
\citep{Schaling-2014} via the R package \texttt{BH}
\citep{Eddelbuettel-2015}. Source code is hosted at
\href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
Installation instructions are available at this homepage, and binary
releases will also be available there.  (Note that {\plant} does not
currently compile on Windows, but this will change soon as R core
upgrade the Windows C++ compiler).

@@

The {\plant} package makes it easier for researchers and forest practitioners
alike to investigate a variety of non-linear demographic phenomena.  Rather
than provide a model with a few entry points, we have developed an extensible
framework that we hope can be used to answer a number of questions at a
variety of scales. The specific use-cases highlighted here arise from our
interest in the assembly of trait mixtures in communities. However a variety
of other uses would be possible, and also extensions to address other aspects
of community assembly. For example,  the physiology of growth could be
modified to account for rising atmospheric CO2 or the effects of competition
for multiple resources could likewise be investigated.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/plant.pdf}
\caption{Physiological dynamics for individual plants vary with size, trait
and light environment. (a) Growth trajectories for plants differing in light
environment and trait (LMA, low = 0.1, solid line; high =  0.267, dashed
line). Lower light environments slow trajectories and interact non-linearly
with traits. (b) Over time, the fraction of living tissue switches from leaf
towards sapwood, with high LMA species having relatively more mass in leaf
than low LMA species. (c) Growth rates peak around 5m high, but this position
varies with both trait and light level (note that this is the derivative of
panel (a)). (d) Growth rates vary with plant size (seedling and mature shown)
and with traits; growth rate differences are more pronounced in high light.
Whole plant light compensation points emerge from the model as the point where
growth rate is zero (x intercept).  The light level used in panels (a) and (c)
is indicated with dots.  See Appendix \ref{sec:plant} for details and code.}
% TODO[Note from James]: [Could you put delta symbols on the y axis rather than d? Also there is alot going in these figures with 4 lines each. You could make it easier for the reader by adding the dashed lines to the keys]
\label{fig:plant}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/patch.pdf}
\caption{Population dynamics for two species competing within a patch. Lines
in panels (a) same growth trajectories for individuals on the
boundary of successive "cohorts" for each species, using a schedule of 
introduction times generated by an adaptive refinement algorithm.
 Red lines are the low LMA
species from figure \ref{fig:plant}, blue lines are the high LMA species.  The
initial growth rate advantage of the low LMA species (see Figure
\ref{fig:plant} c) means it quickly over tops the high LMA species,
suppressing its growth.  Self-thinning leaves space for the high LMA species
to establish below the canopy of the fast species. Panel (b) shows the same
trajectories as in panel (a), but with shading indicating the density $n$ of
individuals at given size and patch age (light = low density, dark = high density).  
(c) The leaf area index at ground
level for each species.  The dotted line is the total leaf area index (sum
across both species). See Appendix \ref{sec:patch} for details and code.}
\label{fig:patch}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/emergent.pdf}
\caption{Emergent size-distributions across a metapopulation.
Grey lines show patterns within individual patches differing in time
since disturbance (age), with blue lines indicating patches of a
specific age. Panels show (a) patterns of abundance, (b) leaf area and
(c) height growth rate against size. Note that units of panels (a) and
(b) are per unit height and per unit ground area, respectively.  See Appendix \ref{sec:emergent} for details and code.}
\label{fig:emergent}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/fitness.pdf}
\caption{Example fitness landscape for LMA, showing potential for stable
coexistence of multiple types.  Fitness is the log of the population growth
rate - at zero species are at equilibrium, above zero they will increase in
density.  (a) Landscape generated by a single species.  At LMA of 0.1 (the
dashed line) there is directional selection for smaller LMA.  At LMA of 0.08
(solid line) there is a branching point (see inset for more detail).  (b)
Landscape generated by two species, holing the first at LMA of 0.08.
Introducing a species at the maximum fitness value from panel (a) reduces the
region of positive fitness considerably (dashed line).  Dotted lines show
subsequent invasions and replacements of the second species, and the solid
line shows a stable fitness landscape where both species exist on a peak in
the fitness landscape.  At this point no further invasion is
possible.  See Appendix \ref{sec:fitness} for details and code.}
\label{fig:fitness}
\end{figure}

@@

\section{Solving for equilibrium seed rain}\label{sec:equilibrium}

@@

\section{Patch level dynamics}\label{sec:patch}

See attached file \url{vignette_patch.R}

\section{Patch level emergent properties}\label{sec:emergent}

See attached file \url{vignette_emergent.R}
